An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-07-06 nips_2016.tex
most proofs done

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}

@@

\subsubsection{Logistic Acceptance Function}\label{ssec:deltas_full}
For our new MH test we use the logistic function: $g(s) =
(1+\exp(-s))^{-1}$. Straightforward arithmetic shows that it satisfies the condition in
Lemma~\ref{lem:detailed_balance}.  While it is slightly less efficient than the Metropolis
test when used on the full dataset, we will see soon that its smoothness allows it to naturally
tolerate substantial variance in its input argument. This in turn will lead to a much more efficient
test on subsets of data.

@@

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$, we
can write
\begin{equation}\label{eq:relationship}
\Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is approximately
normal with variance $\sigma^2(\Delta^*)$. 
But to perform the test in equation (\ref{theexacttest}) we want $\Delta + X$ for
a logistic random $X$ (call it $X_{\rm logistic}$ from now on). 
Our approach is to decompose $X_{\rm logistic}$ as
\begin{equation}\label{eq:deconvolution}
X_{\rm logistic} = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ is a zero-mean Gaussian and $X_{\rm corr}$ is a zero-mean
``correction'' term.  These two add (i.e., their distributions convolve) to form $X_{\rm logistic}$. The acceptance
test is now:
\begin{equation}\label{eq:criteria}
\Delta + X_{\rm logistic} = (\Delta + X_{\rm norm}) + X_{\rm corr} = \Delta^* + X_{\rm corr} >0.
\end{equation}
i.e. given an approximate estimate $\Delta^*$ from the current data minibatch (assuming its
variance is small enough) we test acceptance by adding a random variable $X_{\rm corr}$ and
then accept $\theta'$ if the result is positive.

@@

It is important to understand why we use Equations~\ref{eq:deconvolution} and~\ref{eq:criteria}. The
deconvolution process allows us to use $\Delta' + X_{\rm corr}$ as a proxy for $\Delta + X$. It
works because if we know the distribution\footnote{Computing the distributions of $\varepsilon$ and
$X_{\rm norm}$ reduces to finding their standard deviations.} of $\varepsilon$ (from
Equation~\ref{eq:relationship}), then we can pretend that our $\Delta'$ is in fact $\Delta$. To
``insert'' $\varepsilon$ into our test, we deconvolve $X$ so that one of its components, $X_{\rm
norm}$, plays the part of $\varepsilon$. In light of our revised MH test, the previous discussion
raises the question: in an iteration of MCMC with current $\theta$ and proposed $\theta'$, how do
we know the distribution of $X_{\rm corr}$? We need the correct distribution to sample the
\emph{values} of $X_{\rm corr}$ appropriately. This procedure consists of three major steps.

\begin{enumerate}[noitemsep]
    \item We first need to estimate the standard deviation of $X_{\rm norm}$ (i.e. the standard deviation of $\Delta'$). We use one minibatch data to complete this task. From the proof of Lemma~\ref{lem:gaussian}, we know the randomness comes from $\frac{N}{n}\sum_{i=1}^n \left(\log p(x_i|\theta') - \log p(x_i|\theta) \right)$. Thus, ${\rm var}(\Delta') = \frac{N^2}{n^2}\sum_{i=1}^n{\rm var}\left (  \log p(x_i|\theta') - \log p(x_i|\theta) \right ) = \frac{N^2}{n}{\rm var}\left (  \log p(x_i|\theta') - \log p(x_i|\theta) \right )$. We use data instances in one minibatch to estimate the ${\rm var} \left (  \log p(x_i|\theta') - \log p(x_i|\theta) \right )$. Since there are $n$ data instances in one minibatch, the estimation of ${\rm var} \left (  \log p(x_i|\theta') - \log p(x_i|\theta) \right )$ has high accuracy. In our experiments (see Section~\ref{sec:experiments}) we also use the complete data to obtain the population variance of $\left (  \log p(x_i|\theta') - \log p(x_i|\theta) \right )$, and use the nonparametric Bootstrap to estimate the ${\rm std(std}(\Delta'))$, so that we can construct the confidence interval of ${\rm std}(\Delta')$ to reveal the accuracy of our estimation.

	\item If the ${\rm std}(\Delta') > 1.2$, the decomposition of $X$ (i.e. Equation~\ref{eq:deconvolution}) is unstable. We consume more minibatches to decrease the ${\rm std}(\Delta')$. We update the value of $\Delta'$ by $\Delta' = (k\Delta'_{\rm old} + \Delta'_{\rm new}) / (k+1)$, where $\Delta'_{\rm old}$ is the existing $\Delta'$ which uses $k$ minibatches to computed (usually $k=1$), and $\Delta'_{\rm new}$ is the $\Delta'$ computed from the new minibatch. We also update the value of ${\rm std}(\Delta')$ by ${\rm std}(\Delta') = \sqrt{\left ( k^2 {\rm var}(\Delta'_{\rm old}) + {\rm var}(\Delta'_{\rm new}) \right ) / (k+1)^2}$. Thus, we can decrease the ${\rm std}(\Delta')$ by consuming more data until the ${\rm std}(\Delta') > 1.2$. With help of temperature (see Section~\ref{ssec:discussion}), we will show that usually the mean of minibatch consumed per iteration is below 2 in Section~\ref{sec:experiments}. Later our efficiency analysis in Section~\ref{sec:theory} will reveal that the number of data instance consumed by our algorithm is much fewer than that of existing algorithms.

    \item Once we have ${\rm std}(\Delta')$, we can determine the distribution of $X_{\rm corr}$
    from standard deconvolution techniques. These rely on the well-known fact that the Fourier
    transform of a convolution is the product of the individual Fourier transforms. In practice, we
    pre-compute distributions of $X_{\rm corr}$ for different ${\rm std}(\Delta')$ values to
    facilitate the sampling process.
\end{enumerate}

Algorithm~\ref{alg:our_algorithm} describes our MH test within the MCMC algorithm.

\subsection{Discussions and Practical Considerations of the New MH Test}\label{ssec:discussion}

\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\begin{algorithm}[t]
\Input{Number of samples $T$, minibatch size $m$, pre-computed $X_{\rm
corr}$ distribution, and initial sample $\theta_0$.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta\}$ from $p(\theta)$\;
MAP/ML estimation.}
\For{$t=\{1, \ldots, T\}$}{
    Propose a candidate $\theta'$\ from proposal distribution\;
    Draw a minibatch of $m$ samples, compute $\Delta'(\theta',\theta)$ and ${\rm std}(\Delta')$ for this sample\;
    \While {${\rm std}(\Delta') \ge 1$} {
    	Draw $m$ more samples to augment the current minibatch, and update $\Delta'$ and ${\rm std}(\Delta')$ \;
    }
    Let $\sigma = \sqrt{1 - {\rm std}(\Delta')^2}$, draw a normal variable $X_n \in Norm(0,\sigma)$ and
    $X_c$ from the correction distribution\;

        \eIf{$X_c + X_n + \Delta' > 0$}{
            Accept the candidate, $\theta_{t+1} = \theta'$\;
        }{
            Reject and re-use the old sample, $\theta_{t+1} = \theta$\;
        }

}
\caption{A description of our MH test within the MCMC algorithm.}
\label{alg:our_algorithm}
%\vspace{-10pt}
\end{algorithm}

\textbf{The Convolution}. Figure~\ref{fig:deconvolution} demonstrates
Equation~\ref{eq:deconvolution} and provides three (color-coded) examples of densities for $X_{\rm
norm}$ and $X_{\rm corr}$. The density of $X$, a logistic random variable with mean $\mu = 0$ and
scale $s=1$, is known, as are the $X_{\rm norm}$ densities for different standard deviations. The
deconvolution provides us with the $X_{\rm corr}$ distributions. Note, however, that as the standard
deviation of $X_{\rm norm}$ increases, the $X_{\rm corr}$ distribution becomes increasingly unstable
and ``bumpier'', because the logistic curve has fatter tails than Gaussians. In addition, we cap
the standard deviation possibilities of $X_{\rm norm}$ at $\approx 1.2$ in our experiments. In our
experiments, we discretize the possible standard deviations of $X_{\rm norm}$, and we also limit the
densities to consider the range $[-10,10]$ as shown in Figure~\ref{fig:deconvolution}.

\textbf{Variance Preconditioning}. Returning to the discussion from Section~\ref{sec:introduction},
our test requires a variance check. Equation~\ref{eq:deconvolution} means that the variance of $X$,
which is $\pi^2/3\approx 3.29$ is an upper bound on the variances of $X_{\rm norm}$. Therefore, the
test cannot work if the variance of $X_{\rm norm}$ is too large. The standard deviation of $X$
is about $\sqrt{3.29}\approx 1.81$, so this bounds ${\rm std}(X_{\rm norm})$ (as well as ${\rm
std}(\Delta')$). From the discussion above, we bound the standard deviation at 1.2 so that $X_{\rm
corr}$ can handle the remaining noise. Recall from Section~\ref{ssec:deltas_minibatch} that we
estimate the ${\rm std}(X_{\rm norm})$ each iteration. If the variance/standard deviation test
fails, then we can (a) skip the iteration, (b) increase the minibatch size, (c) change the proposal
to decrease step sizes, or (d) increase temperature, as we now discuss.

\textbf{Temperature}. One way to satisfy the variance precondition is to increase the temperature of
the target distribution. For a temperature $T>1$, the augmented target distribution would become
\begin{equation}\label{eq:log_temperature}
\log p(\theta \mid x_1,\ldots,x_N) \approx \log p(\theta) + \frac{1}{T}\frac{N}{n} \sum_{i=1}^n\log p(x_i \mid \theta).
\end{equation}
with an extra $1/T$ term augmented. As the temperature increases, the values of $\Delta'$ get
smaller in absolute value, thus reducing variance. The flatter posterior results in easier mixing.

\section{Analysis}\label{sec:theory}

In this section, we first perform an error analysis for our new
acceptance test. Second, we follow the implications of this analysis for accuracy of
the estimated target distribution. Third, we compare the speed and accuracy of
our test with prior work. Due to space constraints, all our proofs
are in Appendix~\ref{app:proofs}.

We first introduce some notation. Let our true (full dataset) and
approximate (minibatch) acceptance probabilities be $P_a = g(\Delta) =
\Pr(\Delta + X > 0)$ and $P_a' = \Pr(\Delta' + X_{\rm corr} > 0)$.
Define $X_{\xi} = \varepsilon - X_{\rm norm}$ which means, invoking
Equations~\ref{eq:relationship} and~\ref{eq:deconvolution}, that
$X_\xi$ is the random variable characterizing the accuracy of our
$X_{\rm norm}$.  Thus, $\Delta' = \Delta + X_{\rm norm} + X_\xi$. We
assume the standard deviation of $X_{\rm norm}$ is less than
$\pi/\sqrt{3}$, so that we can use Equation~\ref{eq:deconvolution}.
We denote $F_X(x)$ as the CDF of $X$.

We use some terms from standard Markov chain theory~\cite{Meyn2009}. Let the \emph{transition
kernels} of MCMC on complete and minibatch data at step $i$ be $P_i$ and $P_i'$, respectively. When
these kernels (e.g., $P_i$) are applied on a probability distribution $D_1$, they generate a new
distribution $D_2$, which we write as $P_i \circ D_1 = D_2$. Define $\pi$ to be the stationary
distribution obtained by $P_i$, and $\pi_i'$ to be the stationary distribution obtained by $P_i'$
($\pi'$ needs a subscript in this case).  Finally, we indicate the \emph{total variation distance}
between two kernels using $\| \cdot \|$.

Our first result shows that the distance $\|P_i\circ D - P_i'\circ D\| \le 4\zeta \ell$ for all
distributions $D$.

% Our first result shows that the distance between $P_i$ and $P_i'$ is proportional to our estimation
% error $X_{\xi}$.

\begin{lemma}\label{lem:theory1}
If $|X_\xi| < \zeta$ and $|\nabla F_X(x)| < \ell$ for all $x$, then $\|P_i\circ D-P_i'\circ D\| \le
4\zeta \ell \;\;\forall D$.
\end{lemma}

%% \begin{proof}
%% See Appendix~\ref{app:theory1}.
%% \end{proof}

Thus, the accuracy of our transition kernel $P_i'$ is related to the quality of our estimated
$X_{\rm norm}$. We next show that ${\rm Var}(\Delta')$ is roughly proportional to the step size of
our proposer.

\begin{lemma}\label{lem:theory2}
For one step sampling from $\theta$ to $\theta'$, if the jumping step $\|\theta - \theta'\|_2 <
\epsilon$, and the gradient of the log likelihood is bounded by a constant factor, i.e. $\|\nabla
(\log \Pr(x_i\mid \theta))\|_2 < k$, the variance of $\Delta'$ is bounded by $4\epsilon^2 k^2
(m-\frac{m(m-1)}{N-1})$, where $m$ is the minibatch size, and $N$ is the total data size.
\end{lemma}

%% \begin{proof}
%% See Appendix~\ref{app:theory2}.
%% \end{proof}

Since ${\rm Var}(\Delta') < 4\epsilon^2 k^2 (m - \frac{m(m-1)}{N-1}) $, we can assume the maximum
value of $X_{\xi}$ is proportional to the standard deviation of $\Delta'$, i.e. $\epsilon$. Thus, we
can assume $\zeta=\epsilon C$, where $C$ is a constant factor. We finally introduce
Theorem~\ref{thm:theory3} to characterize the convergence of our samples.

\begin{theorem}\label{thm:theory3}
Assume $\pi$ satisfies $\|P_i \circ D_0 - \pi\| \leq \eta \|D_0 - \pi\|$, where $\eta \in [0, 1)$ for
all distributions $D_0$. Also assume there exists $0 < \rho_t < 1$ such that $\|P'_i \circ D -
\pi_i'\| \leq 2\rho_i \;\; \forall D$. We can get $\| P_t' \circ P_{t-1}' \circ \cdots \circ P_1' \circ D_0 - \pi
\| \leq \sum_{s=1}^t \{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\} \rho_s \alpha_s + \alpha_t $, where
$\alpha_i = \frac{\epsilon_i C }{1-\eta}$ and $\epsilon_i$ is the step size at iteration $i$.
\end{theorem}

%% \begin{proof}
%% See Appendix~\ref{app:theory3}.
%% \end{proof}

Theorem~\ref{thm:theory3} indicates that the difference between the target and sampled distributions
consists of two components: $\rho_i$, which is determined by the efficiency of $P'_i$, and the
$\alpha_i$s, which are determined by the error of our transition kernels. These involve competing
objectives: the larger the step size, the more efficient the $P_i'$ but its error increases.  The
reverse happens with a step size too small.

We explore the efficiency of our minibatch MH test. Theorem~\ref{thm:efficiency} indicates that our minibatch MH test only requires a constant number of instances for each step, however, recall the Lemma~\ref{lem:efficienty_cutting}, the adaptive minibatch MH test requires $O(\sqrt{N})$ number of data instance for each step. Thus, We show that our proposed minibatch MH test is much higher efficient than the known minibatch algorithms.

\begin{theorem}\label{thm:efficiency}
We require $\frac{ C^2 \rho^2 }{\sigma^6 \varepsilon^2}$ minibatch size to ensure $|P_a -P_a'| < \varepsilon$, where $C$ is a positive constant which is smaller than 0.4748, $\sigma^2 $ and $\rho$ is the second and third moment of $(\nabla\log \Pr(x_i|\theta) - \nabla\log\Pr(x_i|\theta')) - E\left[ \nabla\log \Pr(x_i|\theta) - \nabla\log\Pr(x_i|\theta')\right]$, $P_a$ is the accept rate from the complete data, $P_a'$ is the accept rate from minibatch data by our algorithm, $\theta$ is the current parameter, and $\theta'$ is the proposed parameter.

\end{theorem}

>>>>

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}

@@

\subsection{Logistic Acceptance Function}
For our new MH test we use the logistic function:
$g(s)=(1+\exp(-s))^{-1}$. Straightforward arithmetic shows that it
satisfies the condition in Lemma~\ref{lem:detailed_balance}.  While it
is slightly less efficient than the Metropolis test when used on the
full dataset, we will see soon that its smoothness allows it to
naturally tolerate substantial variance in its input argument. This in
turn will lead to a much more efficient test on subsets of data.

@@

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$, we
can write
\begin{equation}\label{eq:relationship}
\Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is approximately
normal with variance $\sigma^2(\Delta^*)$. 
But to perform the test in equation (\ref{theexacttest}) we want $\Delta + X$ for
a logistic random $X$ (call it $X_{\rm log}$ from now on). 
Our approach is to decompose $X_{\rm log}$ as
\begin{equation}\label{eq:deconvolution}
X_{\rm log} = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ is a zero-mean Gaussian and $X_{\rm
  corr}$ is a zero-mean ``correction'' term with distribution $C(X)$.
The two variables are added (i.e., their distributions convolve) to
form $X_{\rm log}$. It will be convenient to express this relationship
in terms of the CDFs $N(X_{\rm norm}) = \Phi(X_{\rm norm}/\sigma)$ and
$S(X_{\rm log})$, where $\Phi(.)$ is the Normal CDF with zero mean and
unit variance and $S(.)$ is the standard logistic function.

Equation (\ref{eq:deconvolution}) implies that $S = N * C$ where $*$ denotes convolution. We will later
derive a numerical representation of $C$ by deconvolution, and using this we can generate
samples of $X_{\rm corr}$.

  The acceptance test is now:
\begin{equation}\label{eq:criteria}
\Delta + X_{\rm log} = (\Delta + X_{\rm norm}) + X_{\rm corr} = \Delta^* + X_{\rm corr} >0.
\end{equation}
i.e. given an estimate $\Delta^*$ from the current data minibatch (assuming its
variance is small enough) we test acceptance by adding a random variable $X_{\rm corr}$ and
then accept $\theta'$ if the result is positive.

@@

%% It is important to understand why we use Equations~\ref{eq:deconvolution} and~\ref{eq:criteria}. The
%% deconvolution process allows us to use $\Delta' + X_{\rm corr}$ as a proxy for $\Delta + X$. It
%% works because if we know the distribution\footnote{Computing the distributions of $\varepsilon$ and
%% $X_{\rm norm}$ reduces to finding their standard deviations.} of $\varepsilon$ (from
%% Equation~\ref{eq:relationship}), then we can pretend that our $\Delta'$ is in fact $\Delta$. To
%% ``insert'' $\varepsilon$ into our test, we deconvolve $X$ so that one of its components, $X_{\rm
%% norm}$, plays the part of $\varepsilon$. In light of our revised MH test, the previous discussion
%% raises the question: in an iteration of MCMC with current $\theta$ and proposed $\theta'$, how do
%% we know the distribution of $X_{\rm corr}$? We need the correct distribution to sample the
%% \emph{values} of $X_{\rm corr}$ appropriately. This procedure consists of three major steps.

%% \begin{enumerate}[noitemsep]
%%     \item We first need to estimate the standard deviation of $X_{\rm norm}$ (i.e. the standard deviation of $\Delta'$). We use one minibatch data to complete this task. From the proof of Lemma~\ref{lem:gaussian}, we know the randomness comes from $\frac{N}{n}\sum_{i=1}^n \left(\log p(x_i|\theta') - \log p(x_i|\theta) \right)$. Thus, ${\rm var}(\Delta') = \frac{N^2}{n^2}\sum_{i=1}^n{\rm var}\left (  \log p(x_i|\theta') - \log p(x_i|\theta) \right ) = \frac{N^2}{n}{\rm var}\left (  \log p(x_i|\theta') - \log p(x_i|\theta) \right )$. We use data instances in one minibatch to estimate the ${\rm var} \left (  \log p(x_i|\theta') - \log p(x_i|\theta) \right )$. Since there are $n$ data instances in one minibatch, the estimation of ${\rm var} \left (  \log p(x_i|\theta') - \log p(x_i|\theta) \right )$ has high accuracy. In our experiments (see Section~\ref{sec:experiments}) we also use the complete data to obtain the population variance of $\left (  \log p(x_i|\theta') - \log p(x_i|\theta) \right )$, and use the nonparametric Bootstrap to estimate the ${\rm std(std}(\Delta'))$, so that we can construct the confidence interval of ${\rm std}(\Delta')$ to reveal the accuracy of our estimation.

%% 	\item If the ${\rm std}(\Delta') > 1.2$, the decomposition of $X$ (i.e. Equation~\ref{eq:deconvolution}) is unstable. We consume more minibatches to decrease the ${\rm std}(\Delta')$. We update the value of $\Delta'$ by $\Delta' = (k\Delta'_{\rm old} + \Delta'_{\rm new}) / (k+1)$, where $\Delta'_{\rm old}$ is the existing $\Delta'$ which uses $k$ minibatches to computed (usually $k=1$), and $\Delta'_{\rm new}$ is the $\Delta'$ computed from the new minibatch. We also update the value of ${\rm std}(\Delta')$ by ${\rm std}(\Delta') = \sqrt{\left ( k^2 {\rm var}(\Delta'_{\rm old}) + {\rm var}(\Delta'_{\rm new}) \right ) / (k+1)^2}$. Thus, we can decrease the ${\rm std}(\Delta')$ by consuming more data until the ${\rm std}(\Delta') > 1.2$. With help of temperature (see Section~\ref{ssec:discussion}), we will show that usually the mean of minibatch consumed per iteration is below 2 in Section~\ref{sec:experiments}. Later our efficiency analysis in Section~\ref{sec:theory} will reveal that the number of data instance consumed by our algorithm is much fewer than that of existing algorithms.

%%     \item Once we have ${\rm std}(\Delta')$, we can determine the distribution of $X_{\rm corr}$
%%     from standard deconvolution techniques. These rely on the well-known fact that the Fourier
%%     transform of a convolution is the product of the individual Fourier transforms. In practice, we
%%     pre-compute distributions of $X_{\rm corr}$ for different ${\rm std}(\Delta')$ values to
%%     facilitate the sampling process.
%% \end{enumerate}

Algorithm~\ref{alg:our_algorithm} describes the M-H iteration using our acceptance test.

\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}

\begin{algorithm}[t]
\Input{Number of samples $T$, minibatch size $m$, pre-computed $X_{\rm
corr}$ distribution, and initial sample $\theta_1$.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta_T\}$ from $p(\theta)$\;}
\For{$t=\{1, \ldots, T\}$}{
    Propose a candidate $\theta'$\ from proposal distribution $q(\theta'\mid\theta_t)$\;
    Draw a minibatch of $m$ points $x_i$, compute $\Delta^*(x,\theta_t,\theta')$ and sample variance $\sigma^2(\Delta^*)$\;
    \While {$\sigma^2(\Delta^*) \ge 1$} {
    	Draw $m$ more samples to augment the current minibatch, and update $\Delta^*$ and $\sigma^2(\Delta^*)$ \;
    }
    Let $\sigma_n^2 = 1 - \sigma^2(\Delta^*)$, draw a normal variable $X_n \in \mathcal{N}(0,\sigma_n^2)$ and
    $X_c$ from the correction distribution\;

        \eIf{$\Delta^* + X_n + X_c > 0$}{
            Accept the candidate, $\theta_{t+1} = \theta'$\;
        }{
            Reject and re-use the old sample, $\theta_{t+1} = \theta_t$\;
        }

}
\caption{A description of M-H sampling with our acceptance test.}
\label{alg:our_algorithm}
%\vspace{-10pt}
\end{algorithm}

\section{Analysis}\label{sec:analysis}

In this section we derive error bounds for our M-H test, and for the approximate target
distribution that it generates. We discuss generation of the correction variable in the
next section. To cut to the chase, it is possible to generate samples $X_{\rm corr}$ with
a CDF error of at most single-precision floating point error, and possibly better. We
therefore treat the $X_{\rm corr}$ variable as a sample from the exact correction distribution
and we will not analyze its errors.

In the most similar prior works, \cite{cutting_mh_2014} uses asymptotic arguments based
on the central limit theorem to argue that its approximate acceptance test error tends to
zero as batch size increases. But no quantitative bounds are given. In
\cite{icml2014c1_bardenet14}, explicit bounds are given, but they depend on a bound on
the quantity:
\begin{equation}
  \max_{1\leq i\leq N}|\log p(x_i|\theta') - \log p(x_i|\theta)|
\end{equation}
while such bounds can be derived easily for simple models, its unclear how to derive them
for a complex model such as a neural network.

In this paper, we rely on quantitave forms of the central limit theorem which rely on measurable
statistics from the minibatch $\{x_i\}$. Thus a sampler using our approach does not need to ``see''
data beyond the current minibatch. This supports use of the sampler on very large datasets, and
in online mode where the dataset is presented as a stream.

The outline of this section is as follows: we first present bounds on the absolute
and relative error (in CDF) of the distribution of $\Delta^*$ vs a Normal distribution. We then show that
these errors bounds are preserved after the addition of other random variables, in particular
$X_n$ and $X_c$. From this it follows that the acceptance test has the same error bound.

The quantitative central-limit result below comes from \cite{explicit-clt05}:

\begin{lemma}\label{lem:quant_clt}
  Let $X_1,\ldots,X_n$ be
a set of zero-mean, independent, identically-distributed random variables with sample mean $\hat{X}$ and
variance $\sigma^2(\hat{X})$ where:
\begin{equation}
  \hat{X} = \frac{1}{n}\sum_{i=1}^nX_i \qquad   \sigma(\hat{X}) = \frac{1}{n}\left(\sum_{i=1}^n(X_i-\hat{X})^2\right)^{\frac{1}{2}}
\end{equation}
then $t=\hat{X}/\sigma(\hat{X})$ has an approximate Student's distribution which approaches a normal
distribution in the limit. Then from \cite{explicit-clt05}:
\begin{equation}\label{eq:clt-bounds}
 \sup_x|{\rm Pr}(t<x) - \Phi(x)| \leq \frac{6.4E|X|^3+2E|X|}{\sqrt{n}}
  \end{equation}
\end{lemma}

So as long as we know the first and third absolute moments $E|X|$ and $E|X|^3$, we can bound the
error of the normal approximation, which decays as $O(n^{-\frac{1}{2}})$. Making the change of variables
$y = x \sigma(\hat{X})$, equation (\ref{eq:clt-bounds}) becomes
\begin{equation}\label{eq:clt-bounds2}
   \sup_y\left|{\rm Pr}(\hat{X}< y) - \Phi\left(\frac{y}{\sigma(\hat{X})}\right)\right| \leq \frac{6.4E|X|^3+2E|X|}{\sqrt{n}}
\end{equation}
which shows that the distribution of $\hat{X}$ approaches the normal distribution $\mathcal{N}(0,\sigma^2(\hat{X}))$
whose variance is $\sigma^2(\hat{X})$, measured from the sample.

To apply this to our test, let $\mu = \frac{1}{N}\sum_{i=1}^N\log\frac{p(x_i|\theta')}{p(x_i|\theta)}$ which is
the expected log data probability ratio across the entire dataset. Then
Let $X_i = N(\log\frac{p(x_i|\theta')}{p(x_i|\theta)}-\mu)$, are zero-mean, IID variables.
Take $n=b$ the minibatch size, and then:
\begin{equation}
  \Delta^*(x,\theta,\theta') = \hat{X} + \Delta(x,\theta,\theta')
\end{equation}
so that $\sigma(\hat{X}) = \sigma(\Delta^*)$.
\begin{corollary}
  We can now substitute into equation (\ref{eq:clt-bounds2}) and displace by the mean, giving:
\begin{equation}\label{eq:clt-bounds3}
   \sup_y\left|{\rm Pr}(\Delta^*< y) - \Phi\left(\frac{y-\Delta}{\sigma(\Delta^*)}\right)\right| \leq \frac{6.4E|X|^3+2E|X|}{\sqrt{b}} = \epsilon(\theta,\theta',b)
\end{equation}
\end{corollary}

We have shown that the distribution of $\Delta^*$ approximates a
Normal distribution with mean $\Delta$ and variance
$\sigma^2(\Delta^*)$, and bounded its error with estimable
quantities. Both $E|X|$ and $E|X|^3$ can be estimated as means of
$|X_i - \hat{X}|$ and $|X_i - \hat{X}|^3$ respectively on each minibatch. We expect this
will often be accurate enough on minibatches with hundreds or
thousands of points, but otherwise bootstrap CIs can be computed from
those sequences. Since the bounds are monotone in $E|X|$ and $E|X|^3$,
using upper bootstrap CI limits will provide high-confidence error bounds.

\begin{lemma} Let $P(x)$ and $Q(x)$ be two cumulative distributions satisfying $\sup_x|P(x)-Q(x)|\leq \epsilon$ with
  $x$ in some real range. Let $R(y)$ be the {\em density} of another
  random variable $y$. Let $P'$ be the convolution $P*R$ and $Q'$ be
  the convolution $Q*R$. Then $P'(z)$ (resp. $Q'(z)$) is the CDF of sum
  $z=x+y$ of independent random variables $x$ with CDF $P(x)$ (resp. $Q(x)$) and y
  with density $R(y)$.  Then
  \begin{equation}
    \sup_x|P'(x)-Q'(x)|\leq \epsilon
  \end{equation}
  \end{lemma}
\begin{proof}
  \begin{equation}
    \begin{split}
      P'(z) - Q'(z) = \int_{-\infty}^{+\infty}(P(z-x)-Q(z-x))R(x) dx
    \end{split}
  \end{equation}
  and since $\sup_x|P(x)-Q(x)|\leq \epsilon$ it follows that for all $z$:
  \begin{equation}
- \epsilon = \int_{-\infty}^{+\infty} -\epsilon R(x) dx \leq \int_{-\infty}^{+\infty}(P(z-x)-Q(z-x))R(x) dx \leq \int_{-\infty}^{+\infty}\epsilon R(x) dx = \epsilon
  \end{equation}
\end{proof}

\begin{corollary}
  If $|{\rm Pr}(\Delta^* < y) - \Phi(\frac{y-\Delta}{\sigma(\Delta^*)})| \leq \epsilon(\theta,\theta',b)$,
  then
  \begin{equation}
    |{\rm Pr}(\Delta^*+X_n+X_c < y) - S(y-\Delta)| \leq \epsilon(\theta,\theta',b)
  \end{equation}
  where $S(x)$ is the standard logistic function, and $X_n$ and $X_c$ are generated as per Algorithm 1. 
  \end{corollary}
%% \subsection{Discussions and Practical Considerations of the New MH Test}\label{ssec:discussion}
\begin{proof}
  Apply the previous lemma twice. First take:
  \begin{equation}
    P(y) = Pr(\Delta^* < y) \qquad {\rm and}\qquad Q(y) = \Phi\left(\frac{y-\Delta}{\sigma(\Delta^*)}\right)
  \end{equation}
  and convolve with the distribution of $X_n$ which has density $\phi(X/\sigma_n)$ where $\sigma_n^2
  = 1 - \sigma^2(\Delta^*)$. This yields the next iteration of $P$ and $Q$
  \begin{equation}
    P'(y) = Pr(\Delta^*+X_n < y) \qquad {\rm and}\qquad Q'(y) = \Phi\left({y-\Delta}\right)
  \end{equation}
  Now we convolve with the distribution of $X_c$ which gives:
  \begin{equation}
    P''(y) = Pr(\Delta^*+X_n+X_c < y) \qquad {\rm and}\qquad Q''(y) = S\left({y-\Delta}\right)
  \end{equation}
  Both steps preserve the error bound $\epsilon(\theta,\theta',b)$
\end{proof}

%% \textbf{The Convolution}. Figure~\ref{fig:deconvolution} demonstrates
%% Equation~\ref{eq:deconvolution} and provides three (color-coded) examples of densities for $X_{\rm
%% norm}$ and $X_{\rm corr}$. The density of $X$, a logistic random variable with mean $\mu = 0$ and
%% scale $s=1$, is known, as are the $X_{\rm norm}$ densities for different standard deviations. The
%% deconvolution provides us with the $X_{\rm corr}$ distributions. Note, however, that as the standard
%% deviation of $X_{\rm norm}$ increases, the $X_{\rm corr}$ distribution becomes increasingly unstable
%% and ``bumpier'', because the logistic curve has fatter tails than Gaussians. In addition, we cap
%% the standard deviation possibilities of $X_{\rm norm}$ at $\approx 1.2$ in our experiments. In our
%% experiments, we discretize the possible standard deviations of $X_{\rm norm}$, and we also limit the
%% densities to consider the range $[-10,10]$ as shown in Figure~\ref{fig:deconvolution}.

%% \textbf{Variance Preconditioning}. Returning to the discussion from Section~\ref{sec:introduction},
%% our test requires a variance check. Equation~\ref{eq:deconvolution} means that the variance of $X$,
%% which is $\pi^2/3\approx 3.29$ is an upper bound on the variances of $X_{\rm norm}$. Therefore, the
%% test cannot work if the variance of $X_{\rm norm}$ is too large. The standard deviation of $X$
%% is about $\sqrt{3.29}\approx 1.81$, so this bounds ${\rm std}(X_{\rm norm})$ (as well as ${\rm
%% std}(\Delta')$). From the discussion above, we bound the standard deviation at 1.2 so that $X_{\rm
%% corr}$ can handle the remaining noise. Recall from Section~\ref{ssec:deltas_minibatch} that we
%% estimate the ${\rm std}(X_{\rm norm})$ each iteration. If the variance/standard deviation test
%% fails, then we can (a) skip the iteration, (b) increase the minibatch size, (c) change the proposal
%% to decrease step sizes, or (d) increase temperature, as we now discuss.

%% \textbf{Temperature}. One way to satisfy the variance precondition is to increase the temperature of
%% the target distribution. For a temperature $T>1$, the augmented target distribution would become
%% \begin{equation}\label{eq:log_temperature}
%% \log p(\theta \mid x_1,\ldots,x_N) \approx \log p(\theta) + \frac{1}{T}\frac{N}{n} \sum_{i=1}^n\log p(x_i \mid \theta).
%% \end{equation}
%% with an extra $1/T$ term augmented. As the temperature increases, the values of $\Delta'$ get
%% smaller in absolute value, thus reducing variance. The flatter posterior results in easier mixing.

%% \section{Analysis}\label{sec:theory}

%% In this section, we first perform an error analysis for our new
%% acceptance test. Second, we follow the implications of this analysis for accuracy of
%% the estimated target distribution. Third, we compare the speed and accuracy of
%% our test with prior work. Due to space constraints, all our proofs
%% are in Appendix~\ref{app:proofs}.
